# Statistical Analysis

In this section, I will lead to systematically analyze each research question using fundamental statistical techniques, including **regression analysis, t-tests, correlation coefficients, and significance tests**. We will uncover patterns in the data and determine whether alcohol consumption impacts students' academic performance and social behaviors.

## **Alcohol Consumption on Grades**  
### **Question**

**Does alcohol consumption during weekdays and weekends significantly affect students' final grades?**  
To investigate this, I will use **linear regression**, with students’ **final grade (G3)** as the dependent variable and **weekday alcohol consumption (Dalc)** and **weekend alcohol consumption (Walc)** as independent variables. This will help determine whether alcohol consumption has a measurable impact on students' final grades.

### **Hypothesis**  
- **Null Hypothesis (H₀):** There is no significant relationship between alcohol consumption and final grades.  
- **Alternative Hypothesis (H₁):** Higher alcohol consumption negatively affects students' final grades.

### **Regression Model Setup**  
```{r, warning=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)

df <- read.csv("student-mat.csv")
model <- lm(G3 ~ Dalc + Walc, data = df)
summary(model)
```
Based on the linear regression model, we obtain the following equation:

\[
G3 = 10.9232 - 0.1862 \times Dalc - 0.1014 \times Walc
\]
where:  

- `Dalc` represents weekday alcohol consumption.  
- `Walc` represents weekend alcohol consumption.  

### **Key Findings**  

1. **Intercept (`10.9232`)**  

   - So we know that if a student does not consume alcohol on weekdays or weekends (`Dalc = 0`, `Walc = 0`), their expected final grade is 10.92.  

2. **Effect of Weekday Alcohol Consumption (`Dalc`)**  

   - Here the estimated coefficient is -0.1862, meaning that for every 1-unit increase in weekday alcohol consumption, the expected final grade decreases.

3. **Effect of Weekend Alcohol Consumption (`Walc`)**  

   - The estimated coefficient is -0.1014, meaning that for every 1-unit increase in weekend alcohol consumption, the expected final grade decreases slightly.  

### **Model Performance**  

- **R² = 0.0035** → The model explains only 0.35% of the variation in final grades.  
- **F-statistic = 0.6803, p-value = 0.507** → Since the p-value is greater than 0.05, we fail to reject the null hypothesis that `Dalc` and `Walc` have any combined effect on `G3`.  

### **Conclusion**  
- Neither weekday nor weekend alcohol consumption has a statistically significant impact on students’ final grades.
- The model does not effectively explain variations in `G3`, meaning other factors like study time, absences, or family support might be more influential in predicting academic performance.  

## **Comparison of Weekday vs Weekend Alcohol Consumption**  
### **Question**

**Are the effects of alcohol consumption different when comparing weekday consumption vs weekend consumption?**  

### **Hypothesis**  
- **Null Hypothesis (H₀):** There is no significant difference between weekday (`Dalc`) and weekend (`Walc`) alcohol consumption.  
- **Alternative Hypothesis (H₁):** There is a significant difference between weekday (`Dalc`) and weekend (`Walc`) alcohol consumption.

### **Paired Two-Sample t-Test Setup**  
```{r}
t_test_result <- t.test(df$Dalc, df$Walc, paired = TRUE)
t_test_result
```
### **Key Findings**  
1. **Mean Difference (`-0.8101`)**  
   - The average alcohol consumption level on weekends (`Walc`) is 0.81 units higher than on weekdays (`Dalc`).  
   - This suggests that students tend to drink significantly more on weekends compared to weekdays.  

2. **p-value Analysis (`< 2.2e-16`)**  
   - Since the p-value is extremely small (`< 0.05`), we reject the null hypothesis.  
   - This indicates a statistically significant difference between weekday and weekend alcohol consumption.  
   - Therefore, students drink more on weekends than they do on weekdays.  

3. **Confidence Interval (`[-0.9074, -0.7129]`)**  
   - The 95% confidence interval for the mean difference does not include zero, reinforcing the statistical significance of the result. The negative values indicate that weekend consumption is consistently higher than weekday consumption.  

### **Model Performance**  

- **t-statistic = -16.379** → We see a large negative value, so it suggests a strong difference between weekday and weekend drinking.  
- **p-value < 2.2e-16** → There is strong evidence against the null hypothesis, confirming a significant difference.  

### **Conclusion**  

- On weekends, students drink a lot more alcohol than they do during the week. The findings imply that students' weekend drinking habits are more prevalent. Therefore, it is reasonable to assume that weekend social events and free time are factors in increased alcohol intake.


## **Correlation Between Social Factors and Alcohol Use**  
### **Question**

**How does students’ tendency to go out with friends correlate with their weekday and weekend alcohol consumption?**  


### **Hypothesis**  
- **Null Hypothesis (`H₀`)**: There is no significant correlation between **going out with friends (`goout`)** and **alcohol consumption (`Dalc` and `Walc`)**.  
- **Alternative Hypothesis (`H₁`)**: There is a significant correlation between **going out with friends (`goout`)** and **alcohol consumption (`Dalc` and `Walc`)**.  

## **Computing Correlation Coefficients**  

```{r}
cor1 <- cor(df$goout, df$Dalc, method = "pearson")
cor2 <- cor(df$goout, df$Walc, method = "pearson")
cor1
cor2
```

### **Key Findings**
1. **Correlation Between goout and Dalc (Weekday Alcohol Use) = 0.267**
  - Although the relationship is not very strong, there is a fairly weak positive correlation, indicating that students who go out more often tend to drink a little more during the week.

2. **Correlation Between goout and Walc (Weekend Alcohol Use) = 0.420 **
  - There is a moderate positive correlation, indicating that students who go out more often drink significantly more on weekends compared to weekdays. The correlation is stronger for weekend drinking, which is what we normaly expect that students socialize and drink more on weekends.
  
### **Conclusion**

- This supports the idea that socializing on the weekends has a greater impact on alcohol consumption than socializing during the week. This brings us to our next investigation, which is whether romantic relationships have an effect on how much alcohol people drink.

## **Predicting Final Grade (`G3`) Using Social and Academic Factors**  
### **Question**

**Can other variables predict students’ final grades?**  

### **Hypothesis**  
- **Null Hypothesis (`H₀`)**: The selected independent variables do not significantly predict students' final grades (`G3`).  
- **Alternative Hypothesis (`H₁`)**: At least one of the selected independent variables significantly predicts students' final grades (`G3`).  

### **Regression Model Setup**  
```{r}
model <- lm(G3 ~ studytime + absences + failures + goout + Walc + Dalc, data = df)
summary(model)
```
Based on the multiple linear regression model, we obtain the following equation:

\[
G3 = 11.4716 + 0.2486 \times studytime + 0.0332 \times absences - 2.1492 \times failures - 0.4477 \times goout + 0.1914 \times Walc - 0.0543 \times Dalc
\]

where:  
- `studytime` represents **weekly study time**.  
- `absences` represents **total school absences**.  
- `failures` represents **number of past class failures**.  
- `goout` represents **going out with friends**.  
- `Walc` represents **weekend alcohol consumption**.  
- `Dalc` represents **weekday alcohol consumption**.  

### **Key Findings:**  

1. **Significant Predictors:**  
   - **`failures (-2.1492, p < 0.001)`:**  
     - The most significant predictor, indicating that each additional past failure decreases `G3` by 2.15 points** on average.  
   - **`goout (-0.4477, p = 0.037)`:**  
     - This is a negative effect, meaning that students who go out more tend to have lower grades.  

2. **Non-Significant Predictors (`p > 0.05`):**  
   - `studytime (p = 0.355)`: It is surprising that there is no strong evidence that weekly study time influences final grades.  
   - `absences (p = 0.223)`: This is also surprising that absence has no significant impact on final grades.  
   - `Walc (p = 0.420)` & `Dalc (p = 0.864)`:
     - Neither weekday nor weekend alcohol consumption significantly predicts `G3`, supporting earlier findings that alcohol consumption does not strongly impact academic performance.  

### **Model Performance:**  

- **R² = 0.1444** → The model explains only 14.44% of the variation in final grades.  
- **F-statistic = 10.91, p-value = 3.07e-11** → The model is statistically significant overall, meaning at least one predictor contributes to explaining `G3`.  

### **Conclusion**  

- The best predictor of final grades is past failures (referred to as "failures"), with each additional failure resulting in an average decrease of 2.15 points.  
- Socializing (going out) has a small but significant negative impact on grades, which suggests that too much social activity may reduce academic performance.  
- The amount of alcohol consumed (`Dalc`, `Walc`) does not have a significant effect on grades*.
- The model accounts for only 14% of the differences in final grades, suggesting that other factors, such as motivation, teaching quality, and home environment, probably have a greater impact on how well students perform. 
 

## **Comprehensive Multiple Regression Analysis**  

### **Question** 

### **Which variables are the most important predictors of students’ final grades?**  

### **Regression Model Setup**  
```{r}
full_model <- lm(G3 ~ ., data = df)
summary(full_model)
```
### **Key Findings:**  

1. **Most Significant Predictors (`p < 0.05`):**  
   - **`G2 (0.9573, p < 2e-16)`:**  
     - The strongest predictor of `G3`, indicating that each unit increase in `G2` leads to an almost 1-unit increase in `G3`.  
   - **`G1 (0.1888, p = 0.0026)`:**  
     - Also a significant predictor, meaning early academic performance is highly predictive of final grades.  
   - **`absences (0.0459, p = 0.0007)`:**  
     - A small but significant positive effect, suggesting that higher absences are slightly associated with increased grades (potentially due to self-study or external factors).  
   - **`famrel (0.3569, p = 0.0019)`:**  
     - We see that better family relationships are positively correlated with higher final grades, emphasizing the role of family support.  

2. **Moderately Significant Predictors (`0.05 < p < 0.1`):**  
   - **`activities (-0.3460, p = 0.0938)`:**  
     - Suggests that participation in extracurricular activities may slightly decrease final grades.  
   - **`age (-0.1733, p = 0.0864)`:**  
     - Indicates that older students tend to have slightly lower grades, which may be due to differences in academic engagement.  

3. **Non-Significant Predictors (`p > 0.1`):**  
   - **Most demographic factors** (e.g., `sex`, `school`, `parent education`, `guardian`, `studytime`, `health`, `alcohol consumption`) were not statistically significant.  
   - **Alcohol consumption (`Dalc`, `Walc`)** remains insignificant, further confirming that drinking behavior does not strongly impact academic performance.


### **Model Performance:**  

- **R² = 0.8458** → The model explains **84.58% of the variation** in final grades, suggesting a very strong fit.  
- **F-statistic = 47.21, p-value < 2.2e-16** → The model as a whole is highly significant, confirming that at least some predictors are meaningful.  


### **Conclusion**  

My analysis shows that a student's final grade (`G3`) is mainly influenced by their earlier grades (`G1`, `G2`). This means that students who do well at the beginning of the year are likely to keep doing well. **Family support (`famrel`)** has a small positive effect on grades, meaning students with good family relationships tend to perform better in school. Absences have a small positive connection to final grades. This suggests that some students can do well even if they skip school, possibly because they study independently. On the other hand, **extracurricular activities** might slightly lower grades, suggesting that it can be hard to balance schoolwork with other activities. Surprisingly, drinking alcohol (`Dalc`, `Walc`) does not appear to influence grades, which suggests that alcohol use does not directly affect academic performance according to this data. The model accounts for about 85% of the differences in final grades, but by removing unneeded factors and looking at how different variables work together, we could improve the predictions even more.


